---
title: "Microglia Morphology Analysis v2.0"
author: Cam
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.width = 9, fig.height = 5.5)

required_pkgs <- c("tidyverse", "rstatix", "lme4", "cluster", "corrplot",
                    "viridis", "spatstat.explore", "ggthemes", "ggpubr",
                    "ggrepel", "factoextra", "pheatmap")
missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing required package(s): ", paste(missing_pkgs, collapse = ", "),
       ".\nInstall with: install.packages(c(",
       paste0('"', missing_pkgs, '"', collapse = ", "), "))")
}
invisible(lapply(required_pkgs, library, character.only = TRUE))

# Palette/theme, metric metadata, data loading, and shared plotting/stats
# helpers all live in data_helpers.R so this report and app.R (the
# interactive Shiny companion) can't drift apart.
source("data_helpers.R")
```

```{r config}

#  Configure analysis. Edit this code chunk

# Which metadata field to compare across everywhere in this report
# (violins, PCA, clustering, mixed models, statistics, spatial plots):
#   "project"  "well_position"  "site"  "wavelength"  "image"
comparison_variable <- "well_position"

# Optional: lock other metadata fields to specific value(s) to zoom in on a
# subset (e.g. compare sites within a single well position). Leave NULL to
# use everything. Example: filter_well_position <- "B07"
filter_project      <- NULL
filter_well_position <- NULL
filter_site          <- NULL
filter_wavelength    <- NULL

# Cells are read from all morphology_results_*.csv files in this folder and
# cached into merged_morphology_data.csv (much faster to reload - see
# Section 2). The cache auto-rebuilds whenever a source CSV is newer than
# it (e.g. new images added); set this to TRUE to force a rebuild anyway
# (e.g. a source CSV was edited in place without its modified time
# changing). sample_metadata.csv is re-applied on every load regardless,
# so editing it never needs a rebuild.
rebuild_data_cache <- FALSE

# Mask non-significant cells in the correlation heatmap (Section 9)
mask_nonsignificant_correlations <- FALSE

# Compute Ripley's K/L in the spatial section (Section 11); can be slow with
# many images / many cells per image
run_ripley <- FALSE

# Save one density-heatmap PNG per image to disk, sorted into a folder per
# well (Section 11); set to FALSE to skip (this writes files as a side
# effect, unlike the rest of the report which only renders into the HTML).
save_density_plots <- FALSE
density_plot_dir   <- "density_plots"

# Save one morphology-comparison PNG per metric to disk, sorted into a
# folder per metric category (Section 6); set to FALSE to skip.
save_morphology_plots <- TRUE
morphology_plot_dir   <- "morphology_plots"
```

## Background and rationale

This pipeline quantifies microglial morphology from single-channel fluorescence images through a transparent, fully scriptable Fiji/ImageJ workflow whose every step rests on established image analysis methods. Cells are segmented by intensity thresholding, touching cells are separated where needed using watershed segmentation (Beucher & Lantuéjoul, 1979), and skeleton branching metrics (branches, junctions, tips, process length) are computed with the AnalyzeSkeleton plugin (Arganda-Carreras et al., 2010). Shape descriptors (area, perimeter, convex hull, circularity, solidity, aspect ratio) and the intensity-weighted polarity measure follow standard morphometric definitions as implemented in ImageJ's particle analysis.

**To run this**: put this `.Rmd` in the same folder as your morphology CSV(s) (the macro's `analysis_csv/` output) and knit. Every CSV whose filename follows the pattern

```
morphology_results_<project>_<well_position>_<site>_<wavelength>_<image>.csv
```

e.g. `morphology_results_MacNets_MGL_LCD_647_B07_s7_w2_204747.csv` - where `B07` is the well position, `s7` is the site/FOV, and `w2` is the wavelength/channel - has its metadata parsed automatically, no metadata file required. Anything preceding those three tags (including tokens like `647` that don't match a recognized tag) is folded into `project`. A `sample_metadata.csv` (columns `sample_id,...`) can still be dropped in the same folder to attach extra annotations.

**To change what's compared**, edit `comparison_variable` in the Configuration chunk above. Every plot, test, and model in this report updates automatically.

```{r do-import, include=FALSE}
# metric_categories, all_cell_metrics, pretty_var, metric_label,
# parse_cell_name, parse_filename_metadata, load_morphology_data,
# aggregate_to_image, compare_metric, compare_all_metrics,
# plot_single_metric_pub, save_morphology_plots_to_disk all come from
# data_helpers.R (sourced in the setup chunk above).

# ---- Load, filter, and set the active comparison variable ---------------
df <- load_morphology_data(".", force_rebuild = rebuild_data_cache)
n_csv_loaded <- attr(df, "n_csv_loaded")

if (!is.null(filter_project))       df <- df %>% filter(project %in% filter_project)
if (!is.null(filter_well_position)) df <- df %>% filter(well_position %in% filter_well_position)
if (!is.null(filter_site))          df <- df %>% filter(site %in% filter_site)
if (!is.null(filter_wavelength))    df <- df %>% filter(wavelength %in% filter_wavelength)
if (nrow(df) == 0)
  stop("filter_* settings in the Configuration chunk removed every row - loosen them.")

valid_vars <- c("project", "well_position", "site", "wavelength", "image")
if (!comparison_variable %in% valid_vars)
  stop("comparison_variable must be one of: ", paste(valid_vars, collapse = ", "))
if (!comparison_variable %in% names(df))
  stop("Column '", comparison_variable, "' isn't present - filename parsing may ",
       "have failed for these files (check they follow the expected naming pattern).")

df$group <- df[[comparison_variable]]
if (all(is.na(df$group)))
  stop("comparison_variable '", comparison_variable, "' is entirely NA for these ",
       "files - check the filenames match the expected pattern, or pick a ",
       "different comparison_variable.")

missing_metrics <- setdiff(all_cell_metrics, names(df))

# Scales plot width to the number of groups being compared so category
# labels on the x-axis don't overlap/squish together (e.g. 5 well positions
# needs more room than 2 genotypes).
morph_plot_width <- max(7, 1.2 + 1.1 * n_distinct(df$group))
```

## 1. Experiment Summary

```{r summary-narrative, echo=FALSE, results='asis'}
n_images  <- n_distinct(df$image_id)
n_cells   <- nrow(df)
n_metrics <- length(intersect(all_cell_metrics, names(df)))
comp_levels <- sort(unique(df[[comparison_variable]]))
comp_label  <- pretty_var(comparison_variable)

other_vars <- setdiff(c("project", "well_position", "site", "wavelength"), comparison_variable)
held <- character(0)
for (v in other_vars) {
  lv <- unique(df[[v]]); lv <- lv[!is.na(lv)]
  if (length(lv) == 1) held <- c(held, paste0(pretty_var(v), " = ", lv))
}
held_txt <- if (length(held) > 0) paste0(" across fixed ", paste(held, collapse = ", ")) else ""

cat(
  "Loaded **", n_csv_loaded, "** CSV file", if (n_csv_loaded != 1) "s" else "",
  " containing **", n_images, "** image", if (n_images != 1) "s" else "",
  " and **", format(n_cells, big.mark = ","), "** microglia.  \n",
  "Comparing **", comp_label, "** (", paste(comp_levels, collapse = ", "), ")", held_txt, ".  \n",
  "**", n_metrics, "** morphology metrics analyzed.\n",
  sep = ""
)
```

## 2. Data Import

The first load parses every source CSV and writes `merged_morphology_data.csv`
next to them; every subsequent load (in this report or in `app.R`) reads that
one merged file instead, which is dramatically faster once there are more
than a handful of source CSVs. The cache is rebuilt automatically whenever a
source CSV is newer than it, or immediately if `rebuild_data_cache <- TRUE`
in the Configuration chunk.

```{r show-import, eval=FALSE, ref.label="do-import"}
```

```{r import-preview}
dplyr::glimpse(df %>% select(source_file, project, well_position, site, wavelength, image,
                             cell_name, image_id, cell_number, all_of(intersect(all_cell_metrics, names(df)))))
```

## 3. Dataset Overview

```{r dataset-overview}
df %>% distinct(source_file, project, well_position, site, wavelength, image) %>%
  arrange(project, well_position, site, wavelength) %>%
  knitr::kable(caption = "One row per source CSV / image, as parsed from filenames")

if (length(missing_metrics) > 0) {
  cat("NOTE: metrics not found in the data (will be skipped throughout):",
      paste(missing_metrics, collapse = ", "), "\n")
} else {
  cat("All", length(all_cell_metrics), "expected metrics are present.\n")
}

tibble(category = names(metric_categories),
       metrics = sapply(metric_categories, paste, collapse = ", ")) %>%
  knitr::kable(caption = "Metric categories")
```

## 4. Quality Control

```{r qc-compute}
img_agg <- aggregate_to_image(df)

qc_bounds <- img_agg %>%
  summarise(q1 = quantile(n_cells, 0.25), q3 = quantile(n_cells, 0.75)) %>%
  mutate(iqr = q3 - q1, low = q1 - 1.5 * iqr, high = q3 + 1.5 * iqr)

img_agg_flagged <- img_agg %>%
  mutate(flag = case_when(
    n_cells < qc_bounds$low  ~ "Low",
    n_cells > qc_bounds$high ~ "High",
    TRUE ~ "OK"
  ))

dup_cells <- df %>% count(cell_name) %>% filter(n > 1)
n_missing_centroid <- sum(is.na(df$centroid_x_um) | is.na(df$centroid_y_um))

qc_summary <- tibble(
  Check = c("CSV files loaded", "Images", "Cells", "Median cells / image",
            "Images flagged: unusually low count", "Images flagged: unusually high count",
            "Duplicate cell_name entries", "Cells missing centroid coordinates",
            "Expected metrics missing from data"),
  Value = c(n_csv_loaded, n_distinct(df$image_id), nrow(df),
            round(median(img_agg$n_cells), 1),
            sum(img_agg_flagged$flag == "Low"),
            sum(img_agg_flagged$flag == "High"),
            nrow(dup_cells), n_missing_centroid, length(missing_metrics))
)
knitr::kable(qc_summary, caption = "Quality-control summary")

if (any(img_agg_flagged$flag != "OK")) {
  img_agg_flagged %>% filter(flag != "OK") %>%
    select(group, image_id, n_cells, flag) %>%
    knitr::kable(caption = "Images flagged for unusual cell counts (Tukey IQR rule)")
} else {
  cat("No images flagged for unusual cell counts.\n")
}

if (nrow(dup_cells) > 0) {
  knitr::kable(dup_cells, caption = "Duplicate cell_name entries (should be unique - investigate)")
}
```

```{r qc-plot}
ggplot(img_agg_flagged, aes(x = reorder(image_id, n_cells), y = n_cells, fill = flag)) +
  geom_col(colour = "black", linewidth = 0.3) +
  coord_flip() +
  scale_fill_manual(values = c(OK = okabe[1], Low = okabe[4], High = okabe[5])) +
  labs(title = "Cells per image", subtitle = "Flagged via Tukey's IQR rule on per-image counts",
       x = NULL, y = "n cells", fill = "QC flag")
```

## 5. Cell Counts

```{r cell-counts}
df %>% count(project, well_position, site, wavelength, name = "n_cells") %>%
  knitr::kable(caption = "Cells per project / well position / site / wavelength")

ggplot(img_agg, aes(x = .data[["group"]], y = n_cells, fill = .data[["group"]])) +
  geom_boxplot(alpha = 0.85, colour = "black", outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.12, alpha = 0.6, size = 1.5, colour = "grey20") +
  labs(title = paste0("Cells per image by ", pretty_var(comparison_variable)),
       x = NULL, y = "n cells", fill = pretty_var(comparison_variable)) +
  theme(legend.position = if (n_distinct(img_agg$group) > 8) "none" else "right")
```

## 6. Morphology Comparisons

Cells pooled within each `comparison_variable` level, one full-size plot per
metric (not faceted, so nothing gets squished): transparent violin
(distribution shape) + boxplot (median/IQR) + jittered points (raw cells).
Every pair of groups is compared directly (Welch's t-test if both pass a
Shapiro-Wilk normality test, otherwise Mann-Whitney U), BH-corrected across
pairs, with significance brackets drawn only for pairs that come out
significant (and only when comparing 8 or fewer groups, so it stays
readable). Plot width scales with the number of groups being compared;
plot height scales with however many significance brackets end up stacked,
so they don't overlap.

```{r morphology-comparisons, results='asis', fig.width=morph_plot_width}
for (cat_name in names(metric_categories)) {
  metrics <- intersect(metric_categories[[cat_name]], names(df))
  if (length(metrics) == 0) next
  cat("\n\n#### ", str_to_title(gsub("_", " ", cat_name)), "\n\n", sep = "")
  for (m in metrics) {
    knitr::opts_chunk$set(fig.height = bracket_plot_height(df, m, "group"))
    print(plot_single_metric_pub(df, m, group_label = pretty_var(comparison_variable)))
  }
}
```

### Saving individual morphology plots

```{r morphology-plots-save}
if (save_morphology_plots) {
  saved <- save_morphology_plots_to_disk(df, morphology_plot_dir, group_label = pretty_var(comparison_variable))
  cat("Saved", length(saved), "morphology plot(s) into", normalizePath(morphology_plot_dir),
      "sorted by metric category.\n")
} else {
  cat("Skipped saving morphology plots (save_morphology_plots = FALSE in Configuration).\n")
}
```

## 7. Statistics

### Pooled cells

Shapiro-Wilk normality per group, then Mann-Whitney/Welch-t (2 groups) or
Kruskal-Wallis/ANOVA (3+), FDR-corrected across the metric panel.

```{r pooled-tests}
pooled <- compare_all_metrics(df)
if (nrow(pooled) > 0) {
  knitr::kable(pooled, digits = 4, caption = "Pooled-cell comparisons, FDR-corrected")
} else {
  cat("Only one", comparison_variable, "level present: add more data to enable comparisons.\n")
}
```

### Image-level aggregation (nesting-corrected)

Cells cut from the same source image aren't independent; this collapses
each image to its median per metric so the statistical unit matches the
true replicate.

```{r aggregated-tests}
knitr::kable(img_agg, digits = 3, caption = "Per-image aggregated values (median of cells)")

agg_tests <- compare_all_metrics(img_agg)
if (nrow(agg_tests) > 0) {
  knitr::kable(agg_tests, digits = 4, caption = "Image-level comparisons, FDR-corrected")
} else {
  cat("Only one", comparison_variable, "level present: add more images to enable comparisons.\n")
}
```

```{r compare-pooled-vs-aggregated}
if (nrow(pooled) > 0 && nrow(agg_tests) > 0) {
  left_join(pooled %>% select(metric, p_pooled = p, fdr_pooled = p_fdr),
            agg_tests %>% select(metric, p_image = p, fdr_image = p_fdr), by = "metric") %>%
    mutate(flag = if_else(fdr_pooled < 0.05 & fdr_image >= 0.05,
                           "Significant only when pooled (likely pseudoreplication)", "")) %>%
    arrange(fdr_pooled) %>%
    knitr::kable(digits = 4, caption = "Pooled vs. image-level significance")
}
```

### Mixed-model sensitivity check

`metric ~ group + (1|image_id)` vs. the null model via likelihood-ratio
test - the proper hierarchical treatment of cells nested in images,
without throwing away data via aggregation.

```{r mixed-models}
n_images_per_group <- df %>% distinct(group, image_id) %>% count(group)
if (nrow(n_images_per_group) >= 2 && all(n_images_per_group$n >= 2)) {
  mm_results <- purrr::map_dfr(all_cell_metrics, function(m) {
    if (!m %in% names(df)) return(NULL)
    f_full <- as.formula(paste(m, "~ group + (1|image_id)"))
    f_null <- as.formula(paste(m, "~ 1 + (1|image_id)"))
    res <- tryCatch({
      m1 <- lme4::lmer(f_full, data = df, REML = FALSE)
      m0 <- lme4::lmer(f_null, data = df, REML = FALSE)
      anova(m0, m1)
    }, error = function(e) NULL)
    if (is.null(res)) return(NULL)
    tibble(metric = m, chisq = res$Chisq[2], df = res$Df[2], p = res$`Pr(>Chisq)`[2])
  })
  mm_results %>% mutate(p_fdr = p.adjust(p, method = "BH")) %>% arrange(p_fdr) %>%
    knitr::kable(digits = 4, caption = "Mixed-model (LRT) p-values, group effect controlling for image")
} else {
  cat("Needs >=2", comparison_variable, "levels with >=2 images each to run mixed models.\n")
}
```

## 8. PCA

```{r pca-setup}
present_metrics <- intersect(all_cell_metrics, names(df))
pca_input_full <- df %>% select(all_of(present_metrics))
keep_rows <- complete.cases(pca_input_full)
pca_input <- pca_input_full[keep_rows, ]

zero_var <- names(pca_input)[sapply(pca_input, function(x) var(x) == 0)]
if (length(zero_var) > 0) {
  cat("Dropping zero-variance metric(s):", paste(zero_var, collapse = ", "), "\n")
  pca_input <- pca_input %>% select(-all_of(zero_var))
}
pca <- prcomp(pca_input, scale. = TRUE)
pca_groups <- df$group[keep_rows]
var_exp <- (pca$sdev^2 / sum(pca$sdev^2)) * 100

cat(nrow(pca_input), "cells (complete cases) across", ncol(pca_input), "metrics.",
    "PC1+PC2 explain", round(sum(var_exp[1:2]), 1), "% of variance.\n")
```

### Scree plot

```{r pca-scree}
factoextra::fviz_eig(pca, addlabels = TRUE, barfill = okabe[1], barcolor = "black",
                      linecolor = okabe[4]) +
  labs(title = "Variance explained per principal component")
```

### PCA scores with confidence ellipses

```{r pca-scores}
scores <- as_tibble(pca$x[, 1:2]) %>% mutate(group = pca_groups)
ggplot(scores, aes(PC1, PC2, fill = group)) +
  geom_point(size = 2.4, alpha = 0.9, shape = 21, colour = "black", stroke = 0.4) +
  stat_ellipse(aes(colour = group), level = 0.68, linewidth = 0.7, show.legend = FALSE) +
  labs(title = "PCA of all morphology metrics",
       x = paste0("PC1 (", round(var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(var_exp[2], 1), "%)"),
       fill = pretty_var(comparison_variable))
```

### Biplot (scores + loadings)

```{r pca-biplot, fig.height=7}
n_grp_pca <- n_distinct(pca_groups)
factoextra::fviz_pca_biplot(pca, habillage = pca_groups, addEllipses = n_grp_pca >= 2,
                             ellipse.level = 0.68, col.var = "black", repel = TRUE,
                             palette = okabe_pal(n_grp_pca), label = "var",
                             title = "PCA biplot (scores + loadings)")
```

### Top contributing metrics

```{r pca-contrib}
factoextra::fviz_contrib(pca, choice = "var", axes = 1:2, top = 10,
                          fill = okabe[1], color = "black") +
  labs(title = "Top contributing metrics (PC1 + PC2)")

as_tibble(pca$rotation[, 1:2], rownames = "metric") %>% arrange(desc(abs(PC1))) %>%
  knitr::kable(digits = 3, caption = "PC1 / PC2 loadings")
```

## 9. Correlation Matrix

```{r correlation, fig.height=8}
cor_mat <- df %>% select(all_of(present_metrics)) %>% cor(use = "pairwise.complete.obs")
strong_pal <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(200)

if (mask_nonsignificant_correlations) {
  p_mat <- corrplot::cor.mtest(df %>% select(all_of(present_metrics)), conf.level = 0.95)$p
  corrplot::corrplot(cor_mat, method = "color", type = "upper", order = "hclust",
                      hclust.method = "ward.D2", col = strong_pal,
                      tl.col = "black", tl.cex = 0.75, tl.srt = 45,
                      p.mat = p_mat, sig.level = 0.05, insig = "pch", pch.cex = 0.8,
                      addCoef.col = "black", number.cex = 0.45,
                      title = "Correlation between morphology metrics (x = not significant, p>=0.05)",
                      mar = c(0, 0, 2, 0))
} else {
  corrplot::corrplot(cor_mat, method = "color", type = "upper", order = "hclust",
                      hclust.method = "ward.D2", col = strong_pal,
                      tl.col = "black", tl.cex = 0.75, tl.srt = 45,
                      addCoef.col = "black", number.cex = 0.45,
                      title = "Correlation between morphology metrics",
                      mar = c(0, 0, 2, 0))
}
```

Metrics are ordered by hierarchical clustering (Ward's method) so
correlated blocks sit adjacent to each other. Set
`mask_nonsignificant_correlations <- TRUE` in the Configuration chunk to
grey out (cross out) pairs that aren't significant at p < 0.05.

## 10. Clustering

```{r clustering-k, fig.height=4}
scaled <- scale(pca_input)
if (nrow(scaled) >= 4) {
  k_range <- 2:min(6, nrow(scaled) - 1)
  sil_width <- sapply(k_range, function(k) cluster::pam(scaled, k)$silinfo$avg.width)
  best_k <- k_range[which.max(sil_width)]

  tibble(k = k_range, avg_silhouette_width = sil_width) %>%
    ggplot(aes(k, avg_silhouette_width)) +
    geom_line(colour = okabe[1], linewidth = 0.8) +
    geom_point(size = 2.5, colour = "black", fill = okabe[1], shape = 21) +
    geom_vline(xintercept = best_k, linetype = "dashed", colour = okabe[4]) +
    scale_x_continuous(breaks = k_range) +
    labs(title = paste0("Optimal k by average silhouette width (k = ", best_k, ")"),
         x = "Number of clusters (k)", y = "Average silhouette width")
} else {
  cat("Need at least 4 cells (complete cases) for clustering.\n")
}
```

```{r clustering-dendro-heatmap, fig.height=7}
if (nrow(scaled) >= 4) {
  pam_fit <- cluster::pam(scaled, best_k)
  hc <- hclust(dist(scaled), method = "ward.D2")

  print(factoextra::fviz_dend(hc, k = best_k, rect = TRUE, cex = 0.55,
                               k_colors = okabe[1:best_k], rect_border = okabe[1:best_k],
                               color_labels_by_k = TRUE,
                               main = "Hierarchical clustering (Ward's method)"))

  rownames(scaled) <- as.character(seq_len(nrow(scaled)))
  pheatmap::pheatmap(scaled,
    annotation_row = data.frame(cluster = factor(pam_fit$clustering),
                                 row.names = rownames(scaled)),
    show_rownames = FALSE, clustering_method = "ward.D2",
    color = colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(100),
    main = "Cluster heatmap (scaled metrics)")
}
```

```{r clustering-composition}
if (nrow(scaled) >= 4) {
  comp_tab <- table(Cluster = pam_fit$clustering, Group = df$group[keep_rows])
  print(knitr::kable(comp_tab, caption = "Cluster membership vs. experimental group"))

  as_tibble(comp_tab) %>%
    ggplot(aes(x = factor(Cluster), y = n, fill = Group)) +
    geom_col(colour = "black", position = "fill") +
    labs(title = "Cluster composition by group", x = "Cluster", y = "Proportion",
         fill = pretty_var(comparison_variable))
}
```

## 11. Spatial Analysis

Uses each cell's centroid (um, in the original-image frame) to describe how
microglia are arranged within each field. Positions come from the macro's
`centroid_x_um` / `centroid_y_um` columns (or approximate crop-corner
coordinates parsed from the filename if those are absent).

**Caveats:** each image is a finite window, so nearest-neighbour and
dispersion statistics are affected by edge effects and by any cells that
were skipped during approval. Treat these as descriptive/exploratory unless
the field boundary and detection completeness are well controlled.

### Centroid maps

```{r spatial-map, fig.height=6}
spatial_metric <- if ("ramification_index_2d" %in% names(df)) "ramification_index_2d" else "area_um2"

df %>%
  ggplot(aes(centroid_x_um, centroid_y_um, colour = .data[[spatial_metric]])) +
  geom_point(size = 2, alpha = 0.85) +
  scale_colour_viridis_c(option = "magma") +
  scale_y_reverse() +
  coord_equal() +
  facet_wrap(~image_id) +
  labs(title = "Cell positions within each image", colour = spatial_metric,
       x = "x (um)", y = "y (um)")
```

### Density heatmaps

```{r spatial-density, fig.height=6}
df %>%
  ggplot(aes(centroid_x_um, centroid_y_um)) +
  stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  geom_point(size = 0.4, alpha = 0.4, colour = "white") +
  scale_y_reverse() + coord_equal() +
  facet_wrap(~image_id) +
  labs(title = "Cell density within each image", x = "x (um)", y = "y (um)", fill = "Density")
```

### Saving per-image density plots by well

Writes one density-heatmap PNG per image to `density_plot_dir/<well_position>/`,
so images from the same well end up in the same folder. Controlled by
`save_density_plots` / `density_plot_dir` in the Configuration chunk.

```{r spatial-density-save}
if (save_density_plots) {
  images <- df %>% distinct(well_position, site, image_id)

  for (i in seq_len(nrow(images))) {
    wp    <- images$well_position[i]
    st    <- images$site[i]
    img   <- images$image_id[i]

    well_label <- if (is.na(wp)) "unknown_well" else wp
    well_dir   <- file.path(density_plot_dir, well_label)
    dir.create(well_dir, showWarnings = FALSE, recursive = TRUE)

    plot_data <- df %>% filter(image_id == img)

    p <- ggplot(plot_data, aes(centroid_x_um, centroid_y_um)) +
      stat_density_2d(aes(fill = after_stat(density)), geom = "raster", contour = FALSE) +
      scale_fill_viridis_c(option = "magma") +
      geom_point(size = 0.6, alpha = 0.5, colour = "white") +
      scale_y_reverse() + coord_equal() +
      labs(title = paste0("Cell density - ", well_label, if (!is.na(st)) paste0(" / ", st) else ""),
           x = "x (um)", y = "y (um)", fill = "Density")

    file_label <- if (!is.na(st)) st else img
    file_label <- str_replace_all(file_label, "[^A-Za-z0-9_-]", "_")
    fname <- file.path(well_dir, paste0(file_label, ".png"))
    ggsave(fname, p, width = 6, height = 6, dpi = 300)
  }

  cat("Saved", nrow(images), "density plot(s) into", normalizePath(density_plot_dir),
      "across", n_distinct(images$well_position), "well folder(s).\n")
} else {
  cat("Skipped saving density plots (save_density_plots = FALSE in Configuration).\n")
}
```

### Nearest-neighbour distance and Clark-Evans dispersion

```{r spatial-nnd}
# Per-image nearest-neighbour distance (NND) and a Clark-Evans-style
# dispersion index R = observed mean NND / expected NND under randomness.
#   R < 1 -> clustered,  R ~ 1 -> random,  R > 1 -> regular/dispersed.
nnd_summary <- df %>%
  group_by(group, image_id) %>%
  filter(dplyr::n() >= 3) %>%
  group_modify(~{
    pts <- as.matrix(.x[, c("centroid_x_um", "centroid_y_um")])
    win <- owin(range(pts[, 1]), range(pts[, 2]))
    pp  <- ppp(pts[, 1], pts[, 2], window = win)
    nnd <- nndist(pp)
    lambda <- pp$n / area.owin(win)
    expected <- 1 / (2 * sqrt(lambda))
    tibble(n_cells = pp$n, mean_nnd_um = mean(nnd), clark_evans_R = mean(nnd) / expected)
  }) %>%
  ungroup()

if (nrow(nnd_summary) > 0) {
  print(knitr::kable(nnd_summary, digits = 3,
               caption = "Per-image nearest neighbor distance and dispersion (R)"))

  if (n_distinct(nnd_summary$group) >= 2) {
    ggplot(nnd_summary, aes(group, clark_evans_R, fill = group)) +
      geom_boxplot(alpha = 0.85, colour = "black", outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.8, colour = "grey20") +
      geom_hline(yintercept = 1, linetype = "dashed", colour = "grey30") +
      labs(title = "Spatial dispersion by group",
           subtitle = "R<1 clustered, R=1 random, R>1 dispersed",
           y = "Clark-Evans R", x = NULL) +
      theme(legend.position = "none")
  }
} else {
  cat("Not enough cells per image (need >=3) for spatial statistics.\n")
}
```

### Ripley's K/L

```{r spatial-ripley, fig.height=7}
if (run_ripley) {
  ripley_data <- df %>%
    group_by(group, image_id) %>%
    filter(dplyr::n() >= 10) %>%
    group_modify(~{
      pts <- as.matrix(.x[, c("centroid_x_um", "centroid_y_um")])
      win <- owin(range(pts[, 1]), range(pts[, 2]))
      pp  <- ppp(pts[, 1], pts[, 2], window = win)
      L <- Lest(pp, correction = "isotropic")
      tibble(r = L$r, L_iso = L$iso, L_theo = L$theo)
    }) %>%
    ungroup()

  if (nrow(ripley_data) > 0) {
    ggplot(ripley_data, aes(r, L_iso - r, colour = group, group = interaction(group, image_id))) +
      geom_line(alpha = 0.7, linewidth = 0.6) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
      facet_wrap(~image_id) +
      labs(title = "Ripley's L-function (L(r) - r)",
           subtitle = "Above 0 = clustering at that scale, below 0 = dispersion",
           x = "r (um)", y = "L(r) - r")
  } else {
    cat("Not enough cells per image (need >=10) for Ripley's K/L.\n")
  }
} else {
  cat("Ripley's K/L skipped (run_ripley = FALSE in Configuration).\n")
}
```

## 12. Appendix

### Session info

```{r session-info}
sessionInfo()
```


